```{r echo=FALSE, eval=TRUE, results='hide'}
source("layout/utils.R")
library(TMB)
library(Matrix)
```

Sparsity {#Sparsity}
========

Large random effect models require sparsity in order to work in TMB.
In this section we will discuss:

- What exactly we mean by *sparsity*.
- How to formulate *sparse* models (the same model can sometimes be formulated as both dense and sparse).
- How to calculate the sparsity pattern of a given TMB model.
- How to visualize sparsity either as a matrix or a graph.
- How to use sparsity for general optimization problems (not just random effects).


## Conditional independence graphs and DAGs

### Conditional independence
There are various graph representations that are commonly used to
visualize probabilistic structure. One such is the *conditional
independence graph*. Say we have a model of four random variables $X_1,...,X_4$ for which the joint density is:

$$p(x_1,x_2,x_3,x_4) \propto f_1(x_1,x_2)f_2(x_2,x_3)f_3(x_3,x_4)f_4(x_4,x_1)$$

The separability of factors on the right hand side implies some conditional independence properties. For instance if $x_1$ and $x_3$ are held constant then $x_2$ and $x_4$ varies independently. We say that $x_2$ and $x_4$ are conditionally independent given $x_1$ and $x_3$.
The conditional independence graph is defined by drawing *undirected* edges between variables occurring in the same factor $f_i$:

```{r, echo=FALSE, eval=TRUE, fig=TRUE}
h <- diag(4)
rownames(h) <- c("X1","X2","X3","X4")
h[1,2] <-1
h[2,3] <-1
h[3,4] <-1
h[4,1] <-1
h <- h + t(h)
plotGraph(h, DAG=FALSE, group=function(x)x==2|x==3)
```

Equivalently the graph may be visualized via its adjacency matrix:

```{r, echo=FALSE, eval=TRUE, fig=TRUE}
hh <- as(forceSymmetric(h),"dsCMatrix")
hh@x[] <- 1
image(hh)
```

This is the **sparsity pattern** of the model.

> The sparsity pattern visualizes the conditional independence structure of the random effects in the model.

### Node removal properties

Important probabilistic properties can be deduced directly from the graph. This is due to the following *node removal* properties.

1. The conditional distribution given node $X_i$ is found by removing $X_i$ and its edges from the graph. For instance conditional on $X_4$ we get the following graph:
    ```{r, echo=FALSE, eval=TRUE, fig=TRUE}
	plotGraph(h[-4,-4], DAG=FALSE, nrow=1)
    ```

2. The marginal distribution wrt. a node $X_i$ is found by removing $X_i$ from the graph and connecting *all* $X_i$'s neighbors. For instance when integrating $X_4$ out of the joint density we get the following graph for the remaining nodes:
    ```{r, echo=FALSE, eval=TRUE, fig=TRUE}
	hm <- h; hm[1,3] <- hm[3,1] <- 1
	hm <- hm[-4, -4]
	plotGraph(hm, DAG=FALSE, nrow=1)
    ```

> Conditioning preserves sparseness. Marginalizing tend to destroy sparseness by adding more edges to the graph.

### Directed acyclic graph

When building models in TMB it is often more natural to specify processes in incremental steps - i.e. through the successive conditional distributions. The previous example could be simulated by drawing the variables $X_1,X_2,X_3,X_4$ one by one in the given order as illustrated by the following *directed* graph:

```{r, echo=FALSE, eval=TRUE, fig=TRUE}
plotGraph(h, DAG=TRUE, group=function(x)x==2|x==3)
```

The graph shows dependencies of any specific node *given past nodes*. The edge from $X_1$ to $X_3$ was not in the original (undirected) graph. This is a so-called **fill-in**.

> Order matters. The DAG is different from the conditional independence graph.

### The effect of adding data

It is convenient to use a box-shape for nodes that represent data. For instance if we pretend that $X_4$ is a data point we would illustrate it by:

```{r, echo=FALSE, eval=TRUE, fig=TRUE}
plotGraph(h, DAG=TRUE, fill_in=FALSE, group=function(x)x!=4, shape=c(rep("circle",3), "square"))
```

Here there are only three variables left. The conditional independence structure of the variables is:

```{r, echo=FALSE, eval=TRUE, fig=TRUE}
plotGraph(hm, DAG=FALSE, nrow=1)
```

which is the same graph as was previously found by integrating $X_4$ out of the joint distribution.

> Data nodes destroy sparsity the same way as marginalization. To avoid this, try to associate each data point with a single random effect.

## The theta logistic example

Consider the ``theta logistic'' population model [@pedersen2011estimation]. This is a
state-space model with state equation

$$u_t = u_{t-1} + r_0\left(1-\left(\frac{\exp(u_{t-1})}{K}\right)^\psi\right) + e_t$$

and observation equation

$$y_t = u_t + v_t$$

where $e_t \sim N(0,Q)$, $v_t \sim N(0,R)$ and $t\in \{0,...,n-1\}$. A
uniform prior is implicitly assigned to $u_0$.

The joint negative log-likelihood of state vector $u$ and measurements
$y$ is implemented in the C++ template `r doxylink("thetalog.cpp")`.
The example can be run by:

```{r, echo=TRUE, eval=FALSE}
runExample("thetalog", exfolder="adcomp/tmb_examples")
```

`r runExample("thetalog", exfolder="../tmb_examples", subarch=FALSE)`

We demonstrate it in the case $n=5$. Here is the DAG

```{r, echo=FALSE, eval=TRUE, fig=TRUE}
n <- 5
parameters <- lapply(c(data, parameters), head, n)
obj <- MakeADFun(data, parameters, random=c("X","Y"), DLL="thetalog")
h <- obj$env$spHess(random=TRUE)
rownames(h) <- c(paste0("Y",1:n), paste0("X", 1:n))
p <- c(1:n + n,1:n)
h <- h[p, p]
plotGraph(h, group=function(x)x>n, shape=c(rep("circle", n), rep("square", n)))
```

This is a standard hidden Markov structure. Each data node is bound to a *single* random effect - hence the data does not introduce additional edges in the random effect structure.

We can use the `image` function from the `Matrix` package to plot the random effect structure (we must first load the Matrix package):

```{R fig=TRUE}
library(Matrix)
obj <- MakeADFun(data, parameters, random=c("X"), DLL="thetalog")
image(obj$env$spHess(random=TRUE))
```

> FIXME: NOT DONE YET !
